### 2015-02-19 doc/thesis/evaluation.tex
Typing LuaFileSystem and adding its evaluation results
Changing the evaluation categories to use percentages

We performed some case studies on existing Lua libraries
to evaluate the applicability of our type system.
For each library, we used Typed Lua to either annotate its modules
or to write statically typed interfaces to its modules through
Typed Lua's description files.
In this chapter we present our evaluation results and discuss some
interesting cases.

We selected seven case studies to evaluate two aspects of Typed Lua:
\begin{enumerate}
\item how precisely it can describe the type of the interface of a module;
\item whether it provides guarantees that the code matches the interface.
\end{enumerate}

@@

LuaSocket \citep{luasocket} was the third case study that we used to
evaluate Typed Lua.
We chose LuaSocket as a case study because it is the most popular Lua library.

There are three case studies that we randomly selected from the LuaRocks
repository, they are: HTTP Digest \citep{luahttpdigest},
Typical \citep{luatypical}, and Mod 11 \citep{luamod11}.
The first one provides client side HTTP digest authentication for Lua.
The second one is an extension to the primitive function \texttt{type}.
The third one is a generator and checker of modulo 11 numbers.
We randomly selected these libraries to evaluate Typed Lua for
annotating existing Lua modules.

The Typed Lua compiler is the last case study that we evaluated.
We chose the Typed Lua compiler as a case study because it is a large application,
and bootstrapping the compiler also provides some confidence in the type system.

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
Lua Standard Library & 80 & 6 & 8 & 35 & 129 & 62\% \\
\hline
MD5 & 13 & 0 & 0 & 0 & 13 & 100\% \\
\hline
LuaSocket & 110 & 1 & 2 & 10 & 123 & 89\% \\
\hline
HTTP Digest & 0 & 0 & 1 & 0 & 1 & 0\% \\
\hline
Typical & 1 & 0 & 0 & 0 & 1 & 100\% \\
\hline
Modulo 11 & 7 & 0 & 0 & 2 & 9 & 78\% \\
\hline
Typed Lua Compiler & 144 & 0 & 1 & 9 & 154 & 94\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for each case study}
\label{tab:evalbycase}
\end{table}

Table \ref{tab:evalbycase} summarizes our evaluation results for each
of the case studies that we used Typed Lua for typing their members.
We split the members of each case study into four categories:
\emph{easy}, \emph{poly}, \emph{inter}, and \emph{hard}.
The \emph{easy} category groups the members that we could give
a precise static type.
The \emph{poly} category groups the members that we made minimal
use of the dynamic type as a replacement for the lack of type parameters.
The \emph{inter} category groups the members that we could give
a static type that is not precise enough because they require
intersection types.
The \emph{hard} category groups the members that we had to rely
on the dynamic type because our type system cannot give a precise
static type to them.
The last column of the table shows the percentage of members that
are under the \emph{easy} category for each case study.
This percentage gives a lower bound on how much static type safety
we could add to each one of our case studies.

In the following sections we discuss each case study in more detail.
For each case study, we split the evaluation results according
to the modules that each one of them include.
This separation by case study allows us to better discuss the
contributions and limitations of our type system.

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
base & 9 & 1 & 2 & 14 & 26 & 35\% \\
\hline
coroutine & 1 & 0 & 0 & 6 & 7 & 14\% \\
\hline
package & 5 & 0 & 0 & 3 & 8 & 62\% \\
\hline
string & 12 & 0 & 0 & 4 & 16 & 75\% \\
\hline
utf8 & 6 & 0 & 0 & 0 & 6 & 100\% \\
\hline
table & 1 & 5 & 0 & 1 & 7 & 14\% \\
\hline
math & 22 & 0 & 4 & 1 & 27 & 81\% \\
\hline
io & 15 & 0 & 0 & 6 & 21 & 71\% \\
\hline
os & 9 & 0 & 2 & 0 & 11 & 82\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Lua Standard Library}
\label{tab:evallsl}
\end{table}

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
md5 & 13 & 0 & 0 & 0 & 13 & 100\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for MD5}
\label{tab:evalmd5}
\end{table}

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
socket & 50 & 0 & 0 & 10 & 60 & 83\% \\
\hline
ftp & 5 & 0 & 1 & 0 & 6 & 83\% \\
\hline
http & 4 & 0 & 1 & 0 & 5 & 80\% \\
\hline
smtp & 7 & 0 & 0 & 0 & 7 & 100\% \\
\hline
mime & 17 & 0 & 0 & 0 & 17 & 100\% \\
\hline
ltn12 & 19 & 1 & 0 & 0 & 20 & 95\% \\
\hline
url & 8 & 0 & 0 & 0 & 8 & 100\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for LuaSocket}
\label{tab:evalsocket}
\end{table}

@@

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
http-digest & 0 & 0 & 1 & 0 & 1 & 0\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for HTTP Digest}
\label{tab:evalhttpdigest}
\end{table}

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
typical & 1 & 0 & 0 & 0 & 1 & 100\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Typical}
\label{tab:evaltypical}
\end{table}

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
mod11 & 7 & 0 & 0 & 2 & 9 & 78\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Modulo 11}
\label{tab:evalmod11}
\end{table}

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
tlast & 46 & 0 & 1 & 0 & 47 & 98\% \\
\hline
tltype & 65 & 0 & 0 & 0 & 65 & 100\% \\
\hline
tlst & 26 & 0 & 0 & 0 & 26 & 100\% \\
\hline
tllexer & 2 & 0 & 0 & 9 & 11 & 18\% \\
\hline
tlparser & 1 & 0 & 0 & 0 & 1 & 100\% \\
\hline
tldparser & 1 & 0 & 0 & 0 & 1 & 100\% \\
\hline
tlchecker & 2 & 0 & 0 & 0 & 2 & 100\% \\
\hline
tlcode & 1 & 0 & 0 & 0 & 1 & 100\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Typed Lua Compiler}
\label{tab:evaltlc}
\end{table}

>>>>

We performed some case studies on existing Lua libraries
to evaluate the design of our type system.
For each library, we used Typed Lua to either annotate its modules
or to write statically typed interfaces to its modules through
Typed Lua's description files.
In this chapter we present our evaluation results and discuss some
interesting cases.

@@

LuaSocket \citep{luasocket} and LuaFileSystem \citep{luafilesystem} were
the third and fourth case studies that we used to evaluate Typed Lua.
We chose them because they are the most popular Lua libraries.
We wrote a script that builds the dependency graph of Lua libraries
that are in the LuaRocks repository, and used this dependency graph
to identify the most popular Lua libraries.

We also randomly selected three case studies from the LuaRocks
repository, they are: HTTP Digest \citep{luahttpdigest},
Typical \citep{luatypical}, and Mod 11 \citep{luamod11}.
The first provides client side HTTP digest authentication for Lua.
The second is an extension to the primitive function \texttt{type}.
The third is a generator and checker of modulo 11 numbers.
We randomly selected three case studies because we wanted to evaluate
Typed Lua for annotating existing libraries that are writting in Lua,
as the previous case studies are mostly libraries that are writting in C.

The Typed Lua compiler is the last case study that we evaluated.
We chose it as a case study because it is a large application.
Besides, it is a case study that evaluates the evolution of a script
to a program.

We used these case studies to evaluate two aspects of Typed Lua:
\begin{enumerate}
\item how precisely it can describe the type of the interface of a module;
\item whether it provides guarantees that the code matches the interface.
\end{enumerate}

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
\hline
Lua Standard Library & 62\% & 5\% & 6\% & 27\% & 129 \\ % 100%
\hline
MD5 & 100\% & 0\% & 0\% & 0\% & 13 \\ % 100%
\hline
LuaSocket & 89\% & 1\% & 2\% & 8\% & 123 \\ % 100%
\hline
LuaFileSystem & 89\% & 0\% & 11\% & 0\% & 19 \\ % 100%
\hline
HTTP Digest & 0\% & 0\% & 100\% & 0\% & 1 \\ % 100%
\hline
Typical & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
\hline
Modulo 11 & 78\% & 0\% & 0\% & 22\% & 9 \\ % 100%
\hline
Typed Lua Compiler & 93\% & 0\% & 1\% & 6\% & 154 \\ % 100%
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for each case study}
\label{tab:evalbycase}
\end{table}

Table \ref{tab:evalbycase} summarizes our evaluation results for each
of the case studies that we used Typed Lua for typing their members.
We split the members of each case study into four categories:
\emph{easy}, \emph{poly}, \emph{over}, and \emph{hard}.
The \emph{easy} category shows the percentage of members that
we could give a precise static type.
The \emph{poly} category shows the percentage of members that
we made minimal use of the dynamic type as a replacement for the
lack of type parameters.
The \emph{over} category shows the percentage of members that
we gave a static type that is not precise enough,
as these members are overloaded functions that require intersection
types to describe their precise static types.
The \emph{hard} category shows the percentage of members that
we had to rely on the dynamic type because they rely on reflection.
The last column of the table shows the total number of members
of each case study.
The results that we obtained for the \emph{easy} category give
a lower bound on how much static type safety
we could add to each one of our case studies.

In the following sections we discuss each case study in more detail.
For each case study, we split the evaluation results according
to the modules that each one of them include,
and we discuss the contributions and limitations that these
evaluation results represent in our type system.

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
\hline
base & 34\% & 4\% & 8\% & 54\% & 26 \\ % 100%
\hline
coroutine & 14\% & 0\% & 0\% & 86\% & 7 \\ % 100%
\hline
package & 62\% & 0\% & 0\% & 38\% & 8 \\ % 100%
\hline
string & 75\% & 0\% & 0\% & 25\% & 16 \\ % 100%
\hline
utf8 & 100\% & 0\% & 0\% & 0\% & 6 \\ % 100%
\hline
table & 15\% & 71\% & 0\% & 14\% & 7 \\ % 100%
\hline
math & 81\% & 0\% & 15\% & 4\% & 27 \\ % 100%
\hline
io & 71\% & 0\% & 0\% & 29\% & 21 \\ % 100%
\hline
os & 82\% & 0\% & 18\% & 0\% & 11 \\ % 100%
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Lua Standard Library}
\label{tab:evallsl}
\end{table}

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
\hline
md5 & 100\% & 0\% & 0\% & 0\% & 13 \\ % 100%
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for MD5}
\label{tab:evalmd5}
\end{table}

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
\hline
socket & 83\% & 0\% & 0\% & 17\% & 60 \\ % 100%
\hline
ftp & 83\% & 0\% & 17\% & 0\% & 6 \\ % 100%
\hline
http & 80\% & 0\% & 20\% & 0\% & 5 \\ % 100%
\hline
smtp & 100\% & 0\% & 0\% & 0\% & 7 \\ % 100%
\hline
mime & 100\% & 0\% & 0\% & 0\% & 17 \\ % 100%
\hline
ltn12 & 95\% & 5\% & 0\% & 0\% & 20 \\ % 100%
\hline
url & 100\% & 0\% & 0\% & 0\% & 8 \\ % 100%
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for LuaSocket}
\label{tab:evalsocket}
\end{table}

@@

\section{LuaFileSystem}

LuaFileSystem is a library that extends the set of functions
for manipulating file systems in Lua.
It contains just the \texttt{lfs} module that is written in C,
so we used Typed Lua's description file to type it.
Table \ref{tab:evallfs} summarizes the evaluation results for LuaFileSystem.

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
\hline
lfs & 89\% & 0\% & 11\% & 0\% & 19 \\ % 100%
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for LuaFileSystem}
\label{tab:evallfs}
\end{table}

Even though we could type most of the functions exported by the
\texttt{lfs} module, we could not type two overloaded functions
due to the lack of intersection types in our type system.

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
\hline
http-digest & 0\% & 0\% & 100\% & 0\% & 1 \\ % 100%
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for HTTP Digest}
\label{tab:evalhttpdigest}
\end{table}

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
\hline
typical & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Typical}
\label{tab:evaltypical}
\end{table}

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
\hline
mod11 & 78\% & 0\% & 0\% & 22\% & 9 \\ % 100%
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Modulo 11}
\label{tab:evalmod11}
\end{table}

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{over} & \textbf{hard} & \textbf{Total} \\
\hline
tlast & 98\% & 0\% & 2\% & 0\% & 47 \\ % 100%
\hline
tltype & 100\% & 0\% & 0\% & 0\% & 65 \\ % 100%
\hline
tlst & 100\% & 0\% & 0\% & 0\% & 26 \\ % 100%
\hline
tllexer & 18\% & 0\% & 0\% & 82\% & 11 \\ % 100%
\hline
tlparser & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
\hline
tldparser & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
\hline
tlchecker & 100\% & 0\% & 0\% & 0\% & 2 \\ % 100%
\hline
tlcode & 100\% & 0\% & 0\% & 0\% & 1 \\ % 100%
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Typed Lua Compiler}
\label{tab:evaltlc}
\end{table}
